Sweat electrolytes and CFTR activity in the era of CFTR modulators

### 2025-04-06 Edith Zemanick/Sweat Variability/sweat_test_analysis.qmd
Fixed sweat test data for Angela

```{r setup}
#| include: false
library(Hmisc)
library(tidyverse)
library(readxl)
library(gtsummary)
knitr::opts_knit$set(root.dir = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Edith Zemanick/Sweat Variability")
```

```{r data cleaning}
#| message: false
# Import data
# source("/Users/timvigers/Library/CloudStorage/Dropbox/GitHub/CF/Edith Zemanick/Sweat Variability/data_import.r")
load("./Data_Cleaned/chco_data.RData")
# Exclude people without demographics
chco <- chco %>% filter(demog_avail == "True")
# Fix diagnoses/exclusion data for CHCO (based on Angela's review)
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude) |
  diagnoses$Code == "Exclude"]
exclude <- exclude[!is.na(exclude)]
chco <- chco[!(chco$patient_id %in% exclude), ]
chco$Diagnosis <- diagnoses$Diagnosis[match(chco$patient_id, diagnoses$PatientID)]
diagnosis_check <- chco %>% select(patient_id, Diagnosis, diagnosis)
# Import NJH data, remove those without demographics
njh <- read_excel("./Data_Raw/SweatVariability2025_CFgenotypes 2024-08-21.xlsx")
# Filter and select similar to CHCO
njh <- njh %>% filter(`Demographics Available`)
# Import detailed diagnoses from Katie
diagnoses <- read.csv("./Data_Cleaned/SweatVariability2025_CFgenotypes.csv",
  na.strings = c("", ".")
)
# Merge NJH and diagnoses
diagnoses <- diagnoses %>% select(Patient.ID, Diagnosis...DataScout)
colnames(diagnoses) <- c("PatientID", "Diagnosis")
# Less detailed diagnoses
diagnoses$Diagnosis <- factor(diagnoses$Diagnosis,
  levels = c(
    "Abnormal sweat choride test without diagnosis of cystic fibrosis",
    "Abnormal sweat test",
    "Alpha-1-antitrypsin deficiency",
    "Alpha-1-antitrypsin deficiency carrier",
    "Alpha-1-antitrypsin deficiency, Positive cystic fibrosis sweat test without diagnosis of cystic fibrosis",
    "CF", "CF carrier, Alpha-1-antitrypsin deficiency",
    "CF, alpha 1-antitrypsin deficiency",
    "CF, alpha 1-antitrypsin deficiency carrier",
    "CF, Alpha-1-antitrypsin deficiency",
    "Cystic fibrosis carrier",
    "Cystic fibrosis carrier, Alpha-1-antitrypsin deficiency",
    "Cystic fibrosis carrier, Alpha-1-antitrypsin deficiency carrier",
    "Cystic fibrosis screening",
    "Cystic fibrosis transmembrane conductance regulator (CFTR)-related disorder",
    "Positive cystic fibrosis sweat test without diagnosis of cystic fibrosis",
    "Primary ciliary dyskinesia",
    "Primary Ciliary dyskinesia, Cystic Fibrosis carrier"
  ),
  labels = c(
    "Non-CF", "Non-CF", "Non-CF", "CF Carrier", "Non-CF", "Cystic Fibrosis",
    "CF Carrier", "Cystic Fibrosis", "Cystic Fibrosis", "Cystic Fibrosis",
    "CF Carrier", "CF Carrier", "CF Carrier", "Non-CF", "CFTR-RD or CRMS",
    "Non-CF", "Non-CF", "CF Carrier"
  )
)
# Patient ids may not match so keep CHCO and NJH separate and check names after
```

```{r}
#| message: false
#| warning: false
df %>%
  ggplot(
    aes(x = Na, y = TotalGenotypeFunction, color = Diagnosis, group = Diagnosis)
  ) +
  geom_point() +
  theme()
df %>%
  ggplot(
    aes(x = Cl, y = TotalGenotypeFunction, color = Diagnosis, group = Diagnosis)
  ) +
  geom_point() +
  theme()
df %>%
  ggplot(
    aes(x = Na / Cl, y = TotalGenotypeFunction, color = Diagnosis, group = Diagnosis)
  ) +
  geom_point() +
  theme()
```

```{r}
df %>%
  filter(Diagnosis == "Cystic Fibrosis") %>%
  mutate(
    `Age group` = cut(Age.at.Test.in.Years, c(-Inf, 18, Inf), right = F),
    `Cl < 60` = Cl < 60
  ) %>%
  select(Na, Cl, TotalGenotypeFunction, `Cl < 60`, `Age group`) %>%
  tbl_summary(by = `Age group`, missing_text = "Missing") %>%
  add_p()
```

>>>>

```{r setup}
#| include: false
library(Hmisc)
library(lubridate)
library(readxl)
library(tidyverse)
library(arsenal)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Edith Zemanick/Sweat Variability",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF/Edith Zemanick/Sweat Variability"
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data cleaning}
#| message: false
# Import CHCO, filter out those with no demo data, rename columns, etc.
chco <- read.csv("./Data_Cleaned/chco_reidentified.csv",
  na.strings = c(
    "", "Test not performed", "Insufficient quant.",
    "repeat sweat chloride testing", "Please reorder testing and re",
    "The received specimen cannot"
  )
)
chco <- chco %>%
  group_by(PatientID) %>%
  filter(Demographics.Available == "TRUE") %>%
  # Get only the earliest test
  slice_min(Age.at.Test.in.Weeks) %>%
  select(
    PatientID, Name, Genotypes1, Genotypes2, Sex, Age.at.Test.in.Years,
    Na, K, Cl, Na_Left, Na_Right, K_Left, K_Right, Cl_Left, Cl_Right
  ) %>%
  rename(Age = Age.at.Test.in.Years) %>%
  mutate(Name = tolower(gsub("[[:punct:] ]+", "", Name)))
# Fix CHCO diagnoses per Angela
chco_diagnoses <- read.csv("./Data_Cleaned/chco_diagnoses.csv")
chco <- full_join(chco, chco_diagnoses, by = join_by(PatientID))
chco$PatientID <- NULL
# For "LESS THAN 10", replace with half the limit of detection
chco[chco == "LESS THAN 10"] <- "5"
chco[, c("Cl", "Cl_Left", "Cl_Right")] <-
  lapply(chco[, c("Cl", "Cl_Left", "Cl_Right")], as.numeric)
# Import NJH and add diagnoses from Katie
njh <- read.csv("./Data_Cleaned/njh_full_list.csv",
  na.strings = c(
    "", "Test not performed", "Insufficient quant.",
    "repeat sweat chloride testing"
  )
)
njh_diagnoses <- read.csv("./Data_Cleaned/njh_diagnoses.csv", na.strings = "")
njh_diagnoses <- njh_diagnoses %>% filter(!is.na(Diagnosis))
njh <- full_join(njh, njh_diagnoses)
# Filter out those with no demo data, rename columns, etc.
njh <- njh %>%
  group_by(Patient.ID) %>%
  filter(Demographics.Available == "TRUE") %>%
  # Get only the earliest test
  slice_min(Age.at.test..in.years.) %>%
  select(
    Patient.ID, Patient.Name, Genotype.1, Genotype.2, Sex, Age.at.test..in.years.,
    Sweat.sodium, Sweat.potassium, Sweat.chloride, Sweat.sodium.left.arm,
    Sweat.sodium.right.arm, Sweat.potassium.left.arm, Sweat.potassium.right.arm,
    Sweat.chloride.left.arm, Sweat.chloride.right.arm, Diagnosis
  ) %>%
  rename(
    Name = Patient.Name, Genotypes1 = Genotype.1,
    Genotypes2 = Genotype.2, Age = Age.at.test..in.years., Na = Sweat.sodium,
    K = Sweat.potassium, Cl = Sweat.chloride, Na_Left = Sweat.sodium.left.arm,
    Na_Right = Sweat.sodium.right.arm, K_Left = Sweat.potassium.left.arm,
    K_Right = Sweat.potassium.right.arm, Cl_Left = Sweat.chloride.left.arm,
    Cl_Right = Sweat.chloride.right.arm
  ) %>%
  mutate(Name = tolower(gsub("[[:punct:] ]+", "", Name)))
njh$Patient.ID <- NULL
# Row bind, fill diagnosis, then remove dupes
df <- rbind(chco, njh)
df <- df %>%
  group_by(Name) %>%
  fill(Diagnosis) %>%
  distinct() %>%
  select(Name, Diagnosis, everything()) %>%
  arrange(Name, Age)
# Use each person's first sweat test
df <- df %>%
  group_by(Name) %>%
  slice_min(Age)
# If there are still duplicates, exclude them
dups <- df$Name[duplicated(df$Name)]
df <- df[!df$Name %in% dups, ]
# For those with left and right, average them
df$Na_mean <- rowMeans(df[, c("Na_Left", "Na_Right")])
df$Na <- coalesce(df$Na, df$Na_mean)
df$Cl_mean <- rowMeans(df[, c("Cl_Left", "Cl_Right")])
df$Cl <- coalesce(df$Cl, df$Cl_mean)
df$K_mean <- rowMeans(df[, c("K_Left", "K_Right")])
df$K <- coalesce(df$K, df$K_mean)
# Get rid of unnecessary columns, remove those with no diagnosis
df <- df %>%
  select(Name, Diagnosis, Genotypes1, Genotypes2, Sex, Age, Na, K, Cl) %>%
  filter(!is.na(Diagnosis))
# Set table options
mycontrols <-
  tableby.control(numeric.stats = c("Nmiss", "meansd", "medianq1q3", "range"))
# Save this for Angela et al. to check
write.csv(df,
  file = "./Data_Cleaned/sweat_variability_analysis_dataset.csv",
  row.names = F, na = ""
)
```

# Data cleaning steps

- Imported re-identified CHCO data from Elin (5,046 rows and 4,748 unique IDs)
- Removed rows with `Demographics Available` not equal to `TRUE` and selected each person's earliest test by age.
  - This resulted in 3,123 rows with 3,084 unique IDs.
- Merged in the diagnoses from Angela.
- Values below the limit of detection were replaced with half the limit (e.g. "LESS THAN 10" was replaced with 5). Approximately 5% of the values were below the L.O.D.

- Imported NJH data (1,958 rows with 1,748 unique IDs) and merged in diagnoses from Katie.
- Removed 1,662 rows with `Demographics Available` not equal to `TRUE` (all of these were `FALSE`)
  - This resulted in 172 rows with unique IDs.
- Concatenated the CHCO and NJH data and removed duplicate rows, resulting in 3,178 rows with 3,162 unique IDs.
- For the remaining people with multiple sweat tests (i.e. tests from both CHCO and NJH), kept only the earliest test (based on age at test). This resulted in 3,169 rows with 3,162 unique IDs.
- The remaining 7 people with different values for the same test in the CHCO and NJH data were excluded. The final analysis dataset includes 3,155 unique participants, 3,111 of which have a diagnosis.
- For people with both left and right arm values, the two were averaged.

# Table 1: Participant Demographics by Diagnosis

```{r results='asis'}
summary(tableby(Diagnosis ~ Sex + Age + Na + K + Cl,
  data = df, control = mycontrols
))
```

# Boxplots

```{r}
ggplot(df, aes(x = Diagnosis, y = Cl)) +
  geom_boxplot() +
  theme_bw()
ggplot(df, aes(x = Diagnosis, y = Na)) +
  geom_boxplot() +
  theme_bw()
ggplot(df, aes(x = Diagnosis, y = Na / Cl)) +
  geom_boxplot() +
  theme_bw()
ggplot(df, aes(x = Diagnosis, y = K)) +
  geom_boxplot() +
  theme_bw()
```
